---
title: "RKNGH analysis"
format: html
---

This is for analysis of the results in the RKNGH-hyperspectral feature importance study.

# Setup

```{r, warning = FALSE, message = FALSE}
# Will be used for easier navigation within the directory structure
here::i_am("r/analysis.qmd")

library(here)
library(rlang)
library(readr)
library(tidyverse)
library(ggthemes)

options(scipen = 9)
knitr::opts_chunk$set(cache = TRUE)
```

```{r}
results <- read_csv(here("results/results.csv"), show_col_types = FALSE) |> 
  mutate(
    # Rename the genes to their proper names, and convert to factor
    gene = case_when(
      gene == "bact" ~ "16S rRNA",
      gene == "cbblr" ~ "cbbLR",
      gene == "fungi" ~ "18S rRNA",
      gene == "phoa" ~ "phoA",
      gene == "urec" ~ "ureC"
    ) |> 
      factor(levels = c("16S rRNA", "cbbLR", "18S rRNA", "phoA", "ureC")),
    # Do the same for the methods
    method = case_when(
      method == "mi" ~ "MI",
      method == "coeffs" ~ "Coeffs",
      method == "cluster" ~ "Cluster",
      method == "perm_imp" ~ "PI",
      method == "ga" ~ "GA",
      method == "consensus" ~ "Consensus (GA)"
    ) |> 
      factor(levels = c("MI", "Coeffs", "Cluster", "PI", "GA", "Consensus (GA)")),
    # Make the RMSE non-negated
    rmse = rmse * -1
  ) |> 
  rename(
    Gene = gene,
    Method = method,
    Waveband = wv,
    RMSE = rmse
  )
```

```{r}
# The data here is calculated with respect to 
data <- read_csv(here("data/data_consol.csv"), show_col_types = FALSE) |> 
  pivot_longer(
    cols = matches("^[0-9]+$"), # Selects all column names which are only integers
    names_to = 'Waveband',
    values_to = 'Intensity'
  ) |> 
  mutate(Waveband = parse_integer(Waveband)) #|>  # Cast wv from char to int
  # group_by(sample, wv) |>
  # summarize(
  #   .groups = 'drop', # Removes the groups after exiting summarize()
  #   mean_o = mean(intensity),
  #   var_o = var(intensity),
  #   iod_o = var_o / mean_o # "iod" = "index of dispersion"
  # )
```

This is for visualizing what region on the spectrum each wavelength is in:

```{r}
vis_light_colors <- binned_scale(
  aesthetics = 'color',
  #scale_name = 'stepsn',
  palette = function(x) c('black', 'purple', 'blue', 'cyan', 'green', 'yellow', 'orange', 'red', 'black'),
  breaks = c(380, 450, 485, 500, 565, 590, 625, 750),
  labels = NULL,
  limits = c(0, 2500),
  show.limits = TRUE,
  guide = 'none'
)
```

# Data distribution

```{r}
data |> 
  select(sample, Waveband, Intensity) |> 
  group_by(sample) |> 
  ggplot() +
  geom_line(aes(x = Waveband, y = Intensity, color = Waveband), linewidth = 0.25) +
  vis_light_colors +
  labs(title = "Data distribution", subtitle = "398 samples overplotted; visible light regions colorized")
```


# RMSE results

```{r}
results |> 
  select(-Waveband) |> 
  unique() |> 
  ggplot() +
  geom_point(aes(x = Method, y = RMSE, fill = Gene), color = "black", size = 3, shape = 21) +
  geom_hline(yintercept = 1) +
  scale_fill_colorblind() +
  labs(title = "Method and gene vs. RMSE")
```

Emphasis here: Method vs. gene effect on RMSE, but easier to compare among genes than faceting


```{r}
results |> 
  select(-Waveband) |> 
  unique() |> 
  group_by(Method) |> 
  summarize(mean_RMSE = mean(RMSE)) |> 
  ggplot() +
  geom_point(aes(x = Method, y = mean_RMSE), size = 3) +
  geom_hline(yintercept = 1) +
  labs(title = "Method vs. mean RMSE among all genes", y = "Mean RMSE")
```

Emphasis above: Overall performance of each method


```{r}
results |> 
  select(-Waveband) |> 
  unique() |> 
  ggplot() +
  geom_point(aes(x = Method, y = RMSE, fill = Gene), color = "black", size = 3, shape = 21) +
  geom_hline(yintercept = 1) +
  scale_fill_colorblind() +
  facet_wrap(~ Gene) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Method and gene vs. RMSE")
```

Emphasis above: different genes just have different results. Some do better, some do worse.

# Waveband selection results

```{r}
results |> 
  select(-RMSE) |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Method, color = Waveband)) +
  facet_wrap(~ Gene) +
  vis_light_colors +
  labs(title = "Wavebands selected by each method, per gene")
```

Emphasis above: Grouping by both gene and method, also showing the disparity in number of wavebands selected

```{r}
results |> 
  select(-RMSE) |> 
  ggplot() +
  geom_histogram(aes(x = Waveband), bins = 210) +
  facet_wrap(~ Gene) +
  labs(title = "Histogram of wavebands selected per gene")

results |> 
  select(-RMSE) |> 
  ggplot() +
  geom_freqpoly(aes(x = Waveband), bins = 210) +
  facet_wrap(~ Gene) +
  labs(title = "Frequency polygon of wavebands selected per gene")
```

Emphasis above: showing the distribution, especially when the points in the plot above overlap






