---
title: "RKNGH analysis"
format: html
---

This is for analysis of the results in the RKNGH-hyperspectral feature importance study.

# Setup

```{r, warning = FALSE, message = FALSE}
# Will be used for easier navigation within the directory structure
here::i_am("r/analysis.qmd")

library(here)
library(rlang)
library(readr)
library(tidyverse)
library(ggthemes)

options(scipen = 9)
knitr::opts_chunk$set(cache = TRUE)
```

Some rows have a `NA` for `Method` and `Waveband`. These represent the baseline models, where no waveband selection methods were applied.

```{r}
results <- read_csv(here("results/results.csv"), show_col_types = FALSE) |> 
  mutate(
    # Rename the genes to their proper names, and convert to factor
    gene = case_when(
      gene == "bact" ~ "16S rRNA",
      gene == "cbblr" ~ "cbbLR",
      gene == "fungi" ~ "18S rRNA",
      gene == "phoa" ~ "phoA",
      gene == "urec" ~ "ureC"
    ) |> 
      factor(levels = c("16S rRNA", "cbbLR", "18S rRNA", "phoA", "ureC")),
    # Do the same for the methods
    method = case_when(
      method == "mi" ~ "MI",
      method == "coeffs" ~ "Coeffs",
      method == "cluster" ~ "Cluster",
      method == "pi" ~ "PI",
      method == "ga" ~ "GA",
      method == "consensus" ~ "Consensus (GA)",
      method == "baseline" ~ "Baseline"
    ) |> 
      factor(levels = c("MI", "Coeffs", "Cluster", "PI", "GA", "Consensus (GA)", "Baseline")),
    # Make the RMSE non-negated
    rmse = rmse * -1,
    region = case_when(
      region == "vis" ~ "Visible only",
      region == "all" ~ "All"
    ) |> 
      factor(levels = c("All", "Visible only"))
  ) |> 
  rename(
    Gene = gene,
    Method = method,
    Waveband = wv,
    RMSE = rmse,
    Region = region,
    R2 = r2,
    MAE = mae,
    Penalty = penalty,
    Ratio = ratio,
    Coefficient = coeff
  ) |> 
  relocate(Gene, Region, Method, Penalty, Ratio, Waveband, Coefficient, RMSE, R2, MAE)
```

```{r}
# The data here is calculated with respect to 
data <- read_csv(here("data/data_consol.csv"), show_col_types = FALSE) |> 
  pivot_longer(
    cols = matches("^[0-9]+$"), # Selects all column names which are only integers
    names_to = 'Waveband',
    values_to = 'Intensity'
  ) |> 
  mutate(Waveband = parse_integer(Waveband))
```

This is for visualizing what region on the spectrum each wavelength is in:

```{r}
vis_light_colors <- binned_scale(
  aesthetics = 'color',
  #scale_name = 'stepsn',
  palette = function(x) c('black', 'purple', 'blue', 'cyan', 'green', 'yellow', 'orange', 'red', 'black'),
  breaks = c(380, 450, 485, 500, 565, 590, 625, 750),
  labels = NULL,
  limits = c(0, 2500),
  show.limits = TRUE,
  guide = 'none'
)
```

# Data distribution

This figure isn't good for the final paper, but it was okay for the preliminary presentation.

```{r}
# data |> 
#   select(sample, Waveband, Intensity) |> 
#   group_by(sample) |> 
#   ggplot() +
#   geom_line(aes(x = Waveband, y = Intensity, color = Waveband), linewidth = 0.25) +
#   vis_light_colors +
#   labs(title = "Data distribution", subtitle = "398 samples overplotted; visible light regions colorized")

# ggsave(here("figures/data_dist.png"), device = "png", dpi = 300)
```


# Prelim figures

Keeping the code for these temporarily as a reference for quicker reproduction

## RMSE results

```{r}
# results |> 
#   select(-Waveband) |> 
#   unique() |> 
#   ggplot() +
#   geom_point(aes(x = Method, y = RMSE, fill = Gene), color = "black", size = 3, shape = 21) +
#   geom_hline(yintercept = 1) +
#   scale_fill_colorblind() +
#   labs(title = "Method and gene vs. RMSE")

# ggsave(here("figures/method_gene_rmse.png"), device = "png", dpi = 300)
```

```{r}
# results |> 
#   select(-Waveband) |> 
#   unique() |> 
#   group_by(Method) |> 
#   summarize(mean_RMSE = mean(RMSE)) |> 
#   ggplot() +
#   geom_point(aes(x = Method, y = mean_RMSE), size = 3) +
#   geom_hline(yintercept = 1) +
#   labs(title = "Method vs. mean RMSE among all genes", y = "Mean RMSE")

# ggsave(here("figures/method_mean_rmse.png"), device = "png", dpi = 300)
```

```{r}
# results |> 
#   select(-Waveband) |> 
#   unique() |> 
#   ggplot() +
#   geom_point(aes(x = Method, y = RMSE, fill = Gene), color = "black", size = 3, shape = 21) +
#   geom_hline(yintercept = 1) +
#   scale_fill_colorblind() +
#   facet_wrap(~ Gene) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Method and gene vs. RMSE")

# ggsave(here("figures/method_gene_rmse_faceted.png"), device = "png", dpi = 300)
```

## Waveband selection results

```{r}
# results |> 
#   select(-RMSE) |> 
#   ggplot() +
#   geom_point(aes(x = Waveband, y = Method, color = Waveband)) +
#   facet_wrap(~ Gene) +
#   vis_light_colors +
#   labs(title = "Wavebands selected by each method, per gene")

# ggsave(here("figures/method_gene_waveband.png"), device = "png", dpi = 300)
```

```{r}
# results |> 
#   select(-RMSE) |> 
#   ggplot() +
#   geom_histogram(aes(x = Waveband), bins = 210) +
#   facet_wrap(~ Gene) +
#   labs(title = "Histogram of wavebands selected per gene")

# ggsave(here("figures/gene_waveband.png"), device = "png", dpi = 300)
```

# Analysis

These are based on the full results after running on SCINet. The main differences with the above are the addition of region (visible light only or all wavebands), R2, and MAE.

## Baseline (all genes)

```{r}
# Look at the baseline scores
results |> 
  filter(Method == "Baseline") |> 
  select(-Method, -Waveband, -Coefficient) |> 
  unique()

results |> 
  filter(Method == "Baseline") |> 
  select(-Method, -Waveband, -Coefficient) |> 
  filter(Gene == '16S rRNA') |> 
  unique()

results |> 
  filter(Method == "Baseline") |> 
  select(-Method, -Waveband, -Coefficient) |> 
  filter(Gene == '18S rRNA') |> 
  unique()
```

```{r}
results |> 
  filter(Method == "Baseline") |> 
  select(-Method, -Waveband, -Coefficient) |> 
  unique() |> 
  ggplot() +
  geom_point(aes(x = Gene, y = RMSE, shape = Region), color = 'red', size = 3) +
  geom_point(aes(x = Gene, y = MAE, shape = Region), color = 'blue', size = 3) +
  geom_point(aes(x = Gene, y = R2, shape = Region), color = 'violet', size = 3) +
  geom_hline(yintercept = 1) +
  geom_hline(yintercept = 0)

results |> 
  filter(Method == "Baseline") |> 
  select(-Method, -Waveband, -Coefficient) |> 
  unique() |> 
  ggplot() +
  geom_point(aes(x = Gene, y = RMSE, shape = Region)) +
  geom_hline(yintercept = 1)
```


Whoa, using visible light only actually *decreased* the error for 16S in comparison with using the entire spectrum. This is a big result for the paper, and it should be emphasized in the manuscript.

18S did very poorly, as the previous paper and this project's preliminary tests suggested. It looks like the hyperspectral-18S pattern just isn't there.

Looks like ureC also did better with vis only, just like 16S. Still underwhelming results there, though. Also underwhelming for phoA and cbbLR.

## 16S

```{r}
results |> 
  filter(Gene == '16S rRNA') |> 
  select(-Waveband, -Coefficient) |> 
  unique()
```


```{r}
results |> 
  filter(Gene == '16S rRNA') |> 
  select(-Waveband) |> 
  ggplot() +
  geom_point(aes(x = Method, y = RMSE, shape = Region)) +
  geom_hline(yintercept = 1)
```

Clustering was totally ineffective for 16S within the visible light region only, and since PI was based on the clustering, it failed too.

It looks like GA had very similar results in both cases, which is really interesting. Maybe they both converged to the same results? Look into this.

For the other methods, the vis only results were pretty conclusively better than the all results.

Also, the coeffs, GA, and consensus vis results were all similar to the baseline, which also needs looking into.

Keep in mind for this whole paper: the consensus method chose far fewer wavebands than the other methods were programmed to or tended to, so naturally the consensus validation model error was a bit higher. The most important thing for consensus is the relative importances assigned via coeffs to the wavebands it chose.

### GA

```{r}
results |> 
  filter(Gene == '16S rRNA' & Method == 'GA') |> 
  select(-Waveband, -Coefficient) |> 
  unique()
```

Yeah, their metrics were very close. The minute differences might just be due to the stochastic algorithm.

```{r}
results |> 
  filter(Gene == '16S rRNA' & Method == 'GA') |>
  select(Region, Waveband) |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Region, color = Waveband)) +
  vis_light_colors

results |> 
  filter(Gene == '16S rRNA' & Method == 'GA') |> 
  filter(Waveband >= 400 & Waveband <= 700) |> 
  select(Region, Waveband) |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Region, color = Waveband)) +
  vis_light_colors

results |> 
  filter(Gene == '16S rRNA' & Method == 'GA') |> 
  filter(Waveband >= 400 & Waveband <= 700) |> 
  select(Region, Waveband, Coefficient) |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Coefficient, color = Region)) +
  geom_hline(yintercept = 0) +
  scale_color_colorblind()
```

This is unexpected though. I thought (hoped) that the GA for all wavebands would have converged to basically the same wavebands as visible only, but that's not the case. I thought maybe the elastic net validation model for vis assigned weights of about 0 to all the picks that didn't line up, but it doesn't look like that's the case either. It might have just been a coincidence, then, although that's a really unsatisfying answer. If there's time later, come back and look into this more.

### Waveband selection results

```{r}
results |> 
  filter(Gene == '16S rRNA') |>
  ggplot() +
  geom_histogram(aes(x = Waveband), binwidth = 10) +
  facet_wrap(~ Region, scales = 'free')

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Waveband >= 400 & Waveband <= 700) |> 
  ggplot() +
  geom_histogram(aes(x = Waveband), binwidth = 10) +
  labs(title = '16S all, but restricted to same range as visible light only') +
  scale_x_continuous(breaks = seq(400, 700, 20))

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  ggplot() +
  geom_histogram(aes(x = Waveband), binwidth = 10) +
  labs(title = '16S vis') +
  scale_x_continuous(breaks = seq(400, 700, 20))
```

Not much agreement here overall. There's some in the ballpark of 580-600, and 670-690 for large peaks, and smaller peaks of agreement around 410, 430-440, 500-510, 540, and 610-620.

But this was WRT all methods, focusing on vis-all differences. Are there any patterns among the methods?

```{r}
results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method == 'Cluster' | Method == 'PI')

results |> 
  filter(Gene == '16S rRNA') |> 
  select(Region, Method, RMSE) |> 
  unique() |> 
  ggplot() +
  geom_point(aes(x = Method, y = RMSE, color = Region), size = 3) +
  scale_color_colorblind() +
  geom_hline(yintercept = 1)

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method != 'Baseline' & Method != 'PI' & Method != 'Cluster') |> 
  ggplot() +
  geom_freqpoly(aes(x = Waveband, color = Method, after_stat(ncount)), binwidth = 10, linewidth = 2, alpha = 0.75) +
  scale_color_colorblind() +
  scale_x_continuous(breaks = seq(400, 700, 10))

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method != 'Baseline' & Method != 'PI' & Method != 'Cluster') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Method, color = Waveband), size = 3, alpha = 0.5) +
  vis_light_colors +
  scale_x_continuous(breaks = seq(400, 700, 10))

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank')

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = 16-rank, color = Waveband), size = 3) +
  scale_x_continuous(breaks = seq(400, 700, 10)) +
  vis_light_colors

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'Visible only') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = scaled_coeff, color = Waveband), size = 3) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(breaks = seq(400, 700, 10)) +
  vis_light_colors
```

Cluster and PI are overlapping pretty hard, but we already knew that since PI is built on clustering results. Wait, it looks like the reason they both did so terribly on vis only is they only chose two wavebands, so of course it would do way worse. But that also indicates that after omitting those two from consideration, visible only outperformed all across the board.

For visible only, MI clearly preferred the red light. Unfortunately, it wasn't among the best methods based on validation results. 

Coeffs, GA, and consensus all thought around 440, exactly 509, around 535, and (exactly) 590 was important. The region from 665-670 was also relatively important to all, but especially MI. Around 600 was important particularly to MI, but also coeffs (not that much for GA or consensus).

As far as the consensus was concerned, the 3 most important wavebands were all orange light. Green and yellow also did okay. Surprisingly, red wasn't that important to that method.

```{r}
results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Method != 'Baseline' & Method != 'PI' & Method != 'Cluster') |> 
  ggplot() +
  geom_freqpoly(aes(x = Waveband, color = Method, after_stat(ncount)), binwidth = 10, linewidth = 2, alpha = 0.75) +
  scale_color_colorblind() +
  scale_x_continuous(breaks = seq(350, 2500, 50))

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Method != 'Baseline' & Method != 'PI' & Method != 'Cluster') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Method, color = Waveband), size = 3, alpha = 0.5) +
  vis_light_colors +
  scale_x_continuous(breaks = seq(350, 2500, 50))

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank')

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = 16-rank, color = Waveband), size = 3) +
  scale_x_continuous(breaks = seq(350, 2500, 50)) +
  vis_light_colors

results |> 
  filter(Gene == '16S rRNA') |> 
  filter(Region == 'All') |> 
  filter(Method == 'Consensus (GA)') |> 
  select(Waveband, Coefficient) |> 
  mutate(
    abs_coeff = abs(Coefficient),
    scaled_coeff = Coefficient / max(Coefficient),
    scaled_abs_coeff = abs_coeff / max(abs_coeff)) |> 
  arrange(abs_coeff |> desc()) |> 
  rowid_to_column(var = 'rank') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = scaled_coeff, color = Waveband), size = 3) +
  geom_hline(yintercept = 0) +
  scale_x_continuous(breaks = seq(350, 2500, 50)) +
  vis_light_colors
```

When considering all, the red focus shifts to around 725 (all methods). But there's a good amount of energy from MI, GA, and coeff around 1450, and there's a very large spike from around 1875-1975 with all four methods contributing some.

The consensus method had a strong preference for infrared wavebands, especially around 1900-1950.

When comparing the magnitude of consensus validation model coefficients between vis and all, vis had a much larger magnitude of its coeffs. I'm not sure to what extent that matters here, though. 

Update: I thought about it, and I think that's just a function of the number of variables in each model. If both models (all and vis) are trying to predict values in the ballpark of 10, for instance, vis has way fewer wavebands, so they'll have overall higher magnitudes. And vice versa for all. Even though y was scaled (Check this if putting this point in paper or presentation!), the same argument applies.

### Baseline (16S)

```{r}
results |> 
  filter(Method == 'Baseline') |> 
  filter(Gene == '16S rRNA') |> 
  ggplot() +
  geom_point(aes(x = Waveband, y = Coefficient, color = Waveband)) +
  geom_hline(yintercept = 0) +
  vis_light_colors +
  facet_wrap(~ Region, scales = 'free') +
  labs(title = '16S baseline coefficients')

results |> 
  filter(Method == 'Baseline') |> 
  filter(Gene == '16S rRNA') |> 
  ggplot() +
  geom_line(aes(x = Waveband, y = Coefficient, color = Waveband)) +
  geom_hline(yintercept = 0) +
  vis_light_colors +
  facet_wrap(~ Region, scales = 'free') +
  labs(title = '16S baseline coefficients')
```

Looking at the baseline model coefficients for 16S, both all and vis only, since these were among the best overall (and most important since they're for 16S). 16S vis baseline is the most important overall.

One of the key patterns to look for is regions with high oscillation vs pretty steady changes. For instance, look at the red light in vis only for a steady rate of change, or the middle IR section in all. But one interesting pattern going on is that the wavebands with the highest coefficients are in high oscillation areas. I'll need to think on the implications there. Especially in vis light for the purples, what does it mean when one of the highest peaks is right next to the global min?

## 18S

```{r}
results |> 
  filter(Method == 'Baseline') |> 
  filter(Gene == '16S rRNA' | Gene == '18S rRNA') |> 
  ggplot() +
  geom_line(aes(x = Waveband, y = Coefficient, color = Waveband)) +
  geom_hline(yintercept = 0) +
  vis_light_colors +
  facet_wrap(Gene ~ Region, scales = 'free') +
  labs(title = '16S and 18S baseline coefficients')
```

Wait, I just wanted to compare some good baseline models to bad baseline models. If 18S baseline looks like this, maybe that explains why it's so bad? It looks like it chose L1 penalty over L2.

```{r}
results |> 
  filter(Method == 'Baseline') |> 
  filter(Gene == '18S rRNA' | Gene == '16S rRNA') |> 
  select(Region, Gene, Penalty, Ratio) |> 
  unique()

results |> 
  filter(Method == 'Baseline') |> 
  filter(Gene == '18S rRNA') |> 
  select(Region, Waveband, Coefficient) |> 
  mutate(
    zero_coeff = near(Coefficient, 0.0)
  ) |> 
  group_by(Region) |> 
  count(zero_coeff)
```

Yeah the vast majority of the baseline models for 18S had zero coefficients. But they went through the exact same model training + tuning process as the other ones, and this gave the best results for cross validation, so it's still valid. It's always possible that 18S's poor overall performance is just due to random chance with this train/test split, but that's always the case for any ML thing like this.

```{r}
results |> 
  filter(Gene == 'cbbLR') |> 
  ggplot() +
  geom_hline(yintercept = 0) +
  geom_line(aes(x = Waveband, y = Coefficient, color = Waveband)) +
  facet_wrap(~Region + Method, scales = 'free') +
  vis_light_colors
```

```{r}
results |> 
  group_by(Gene, Region, Method) |> 
  count()
```











